Sample Clustering

### 2017-04-04 Clustering.Rmd
Adding new analyses

getPrecision<-function(actual,predicted) {
  retrieved <- sum(predicted==1)
  precision <- sum(predicted==1 & actual==1) / retrieved
  precision
}
getRecall<-function(actual,predicted) {
  retrieved <- sum(predicted==1)
  recall <- sum(predicted==1 & actual==1) / sum(actual==1)
  recall
}

@@

patient.info = read.patient.info(patient.file)

patient.info$Patient = gsub("/", "_", patient.info$Patient)
#head(patient.info)

@@

sum.patient.data = summarise.patient.info(patient.info)
sum.patient.data = as.data.frame(subset(sum.patient.data, Initial.Analysis)) ## For now
pander(sum.patient.data)
```

@@

plot.pca<-function(pt) {

  info = arrange(patient.data[[pt]]$info, Endoscopy.Year)

  plots = list()
  for (sv in c('seg.vals','norm.seg.vals')) {

    x1 = patient.data[[pt]][[sv]][, patient.data[[pt]]$info$Samplename]
    x1 = x1[,patient.data[[pt]]$info$Samplename]
    colnames(x1) = paste(patient.data[[pt]]$info$Endoscopy.Year, '(', patient.data[[pt]]$info$Pathology, ')', sep='' )

@@

pplots = lapply(subset(sum.patient.data, Status == 'P')$Patient, plot.pca)
npplots = lapply(subset(sum.patient.data, Status == 'NP')$Patient, plot.pca)
```

```{r echo=F, message=F, warning=F, pcaplots, fig.align='left', fig.height=30, fig.width=20}
grid.arrange(grobs=pplots, nrow=length(subset(sum.patient.data, Status == 'P')$Patient), top="Progressors")
grid.arrange(grobs=npplots, nrow=length(subset(sum.patient.data, Status == 'NP')$Patient), top="Non-Progressors")
```

@@

#### EM
```{r clust, echo=F, message=F, warning=F, fig.height=4, fig.width=4}

em.dr<-function(patient, status, x1, var.regions) {
  cat(paste('**', patient, '**'))
  clusters=c(2:4)

@@

  for (g in clusters) {
    model = densityMclust(x1, G=g)
    highCluster = which.max(colSums(model$parameters$mean))

    tryCatch({
      mod1dr = MclustDR(model, lambda = 0.5)

      #     par(mfrow=c(2,1)) 
      plot(mod1dr, what="evalues")
      title(paste(pt, '(',status,') G=', g, sep=''))
      plot(mod1dr, what='classification') # while this looks like PCA, it's not quite
      title(paste(pt, '(',status,') G=', g, sep=''))
      #     dev.off()
      plot(mod1dr, what='density', dimens=1)
      title(paste(pt, '  G=', g, sep=''))
    }, error = function(e) {
      cat(paste(pt, g, "failed DR"))
    })

    # a1 = rownames(x1[which(model$classification == highCluster),] )
    # if (nrow(var.regions) > 0) {
    #   am = merge(a1, var.regions, all=T)
    #   pander( as.data.frame(rownames(x1[which(model$classification == highCluster),])), caption=paste("From the cluster with the highest means the regions above are selected, of these ",length(setdiff(rownames(var.regions), a1))," regions are not found in the variable regions selected using cutoffs.", length(intersect(rownames(var.regions), a1)), " are in both."))
    # } else {
    #   pander(as.data.frame(a1),caption=paste("From the cluster with the highest means the regions above are selected. This patient had no variable regions selected based on cutoffs"))
    # 
    # }
    llk[[as.character(g)]] = model
  }
  return(llk)
}

llk = as.data.frame(matrix(nrow=nrow(sum.patient.data),ncol=3,dimnames=list(c(sum.patient.data$Patient), c(2:4))))
for (pt in sum.patient.data$Patient) {
    status = subset(sum.patient.data, Patient == pt)$Status

@@

```{r kmeansraw, echo=F, message=F, warning=F, eval=F, fig.height=6, fig.width=6} 
for (pt in sum.patient.data$Patient) {
    status = subset(sum.patient.data, Patient == pt)$Status
    x1 = patient.data[[pt]]$seg.vals
    rownames(x1) = apply(patient.data[[pt]]$seg.vals[,c(1:4)], 1, function(x) strip.whitespace(paste(x[['chrom']], ':', x[['start.pos']], '-', x[['end.pos']], sep=''))) 
    x1 = x1[,patient.data[[pt]]$info$Samplename]

@@

```{r clust2, echo=F, message=F, warning=F, fig.height=4, fig.width=4}
llk = as.data.frame(matrix(nrow=nrow(sum.patient.data),ncol=3,dimnames=list(c(sum.patient.data$Patient), c(2:4))))
for (pt in sum.patient.data$Patient) {
    status = subset(sum.patient.data, Patient == pt)$Status
    x1 = patient.data[[pt]]$norm.seg.vals
    rownames(x1) = apply(patient.data[[pt]]$seg.vals[,c(1:4)], 1, function(x) strip.whitespace(paste(x[['chrom']], ':', x[['start.pos']], '-', x[['end.pos']], sep=''))) 
    x1 = x1[,patient.data[[pt]]$info$Samplename]

@@

# Log liklihood doesn't separate them individually or the range
llk.diff = apply(llk, 1, function(x) diff(range(x)) )
#wilcox.test(llk.diff[subset(sum.patient.data, Status == 'NP')$Patient], llk.diff[subset(sum.patient.data, Status == 'P')$Patient])

@@

#### Kmeans
```{r kmeansnorm, echo=F, message=F, warning=F, fig.height=6, fig.width=6} 
for (pt in sum.patient.data$Patient) {
    status = subset(sum.patient.data, Patient == pt)$Status
    x1 = patient.data[[pt]]$norm.seg.vals
    rownames(x1) = apply(patient.data[[pt]]$seg.vals[,c(1:4)], 1, function(x) strip.whitespace(paste(x[['chrom']], ':', x[['start.pos']], '-', x[['end.pos']], sep=''))) 
    x1 = x1[,patient.data[[pt]]$info$Samplename]

@@

apclust.data<-function(segdata, samples) {
  x1 = segdata[,intersect(colnames(segdata), samples)]
  x1 = x1[, samples]
  rownames(x1) = (segdata[,c(1:4)] %>%
    rowwise() %>%
    mutate(location=paste(paste(chrom, arm, sep=''), '.', start.pos, '-', end.pos, sep='')))$location
  return(x1)  
}

@@

```{r apclust, echo=F, message=F, warning=F, fig.height=10, fig.width=10, fig.align='left'}
q=0.2

@@

## Using the segmentation values
sum.patient.data$years = with(sum.patient.data, paste(start.year, end.year, sep='-'))
sim = sum.patient.data[,c('Status','years')]
rownames(sim) = sum.patient.data[,'Patient']
for (pt in sum.patient.data$Patient) {
  patient.data[[pt]]$info = arrange(patient.data[[pt]]$info, Endoscopy.Year, desc(Pathology))

@@

  last = ncol(x1)
  if (patient.data[[pt]]$info$Status[1] == 'P' & pt != 'AD0252')
    last = sort(grep('HGD|IMC', patient.data[[pt]]$info$Pathology), decreasing=T)[1]
  if (ncol(x1) <= 15) {
    plot(ac, x1[,c(c(1:ncol(x1))[-last], last)], main=pt)
  } else {
    plot(ac, x1[,c(1, sort(sample(c(1:(ncol(x1)-1))[c(-1,-last)], 8)) ,last) ], main=pt)
  }

@@

# May be a difference
ttps = t.test( subset(sim, Status == 'NP')$ss, subset(sim, Status == 'P')$ss )$p.value
ttpn = t.test( subset(sim, Status == 'NP')$ns, subset(sim, Status == 'P')$ns )$p.value
ttpc = t.test( subset(sim, Status == 'NP')$clusters, subset(sim, Status == 'P')$clusters )$p.value

@@

g2 = ggplot(sim[,c('Status','clusters')], aes(y=clusters, x=Status, fill=Status)) + geom_boxplot() + 
  geom_jitter(width=0.2) + stat_boxplot(geom ='errorbar') +
  labs(title=paste("p-value", signif(ttpc, 3)), y='num. clusters')
```

@@

```{r apglmclus, echo=F, message=F, warning=F}
fit = glm(Status~clusters, sim, family=binomial(link='logit'))
#pander(fit, justify='left')

sim = sim[, c('Status', 'clusters', 'ns','ss')]
actualMatrix = matrix(0,1000,2,dimnames=list(c(), c('prec','recall')))
for (i in 1:nrow(actualMatrix)) {
  trainingRows = sample(1:nrow(sim), nrow(sim)-5)
  train = sim[trainingRows,]
  test = sim[-trainingRows,]

@@

  actual = test$Status == 'P'
  predicted = pred>0

  actualMatrix[i,'prec'] = getPrecision( actual, predicted )
  actualMatrix[i, 'recall'] = getRecall( actual, predicted )
}

# ggplot(as.data.frame(actualMatrix), aes(x=recall, y=prec)) + geom_point() +
#   geom_hline(yintercept = 0.5, col='red', alpha=0.5) +
#   labs(title='Precision/Recall for Progressor/NP prediction')

@@

cols = c('chr','cluster','P.x', 'x')
hgmx = matrix(nrow=0,ncol=length(cols),dimnames=list(c(), cols))
for (chr in c(1:22)) {
  for (i in 1:length(ac@clusters)) {
    cl = x1[ac@clusters[[i]],]

@@

```{r hgmx, echo=F, message=F, warning=F, fig.height=6, fig.width=6}
ggplot(subset(hgmx, x > 0), aes(x=chr,y=P.x,color=factor(cluster))) + ylim(0,0.5)+ 
  geom_point() + 
  geom_line(alpha=0.1) + 
  labs(title='Probability of cluster assignment', x='Chromosome', y='P(x)')
```

@@

@@

```{r xval_pt, echo=F, warning=F, message=F, fig.align='left', fig.height=4, fig.width=4}
# Leave out patients
loo_xval = matrix(ncol=3, nrow=0, dimnames=list(c(), c('net.similarity', 'sum.similarity', 'clusters')))
for (i in 1:nrow(sum.patient.data)) {
  simL = sum.patient.data[-i,c(2:4)]
  rownames(simL) = sum.patient.data[-i,1]

@@

  for (pt in sum.patient.data$Patient) {
      x1 = apclust.data(patient.data[[pt]]$seg.vals, patient.data[[pt]]$info$Samplename[-sample(1:(length(patient.data[[pt]]$info$Samplename)), 1)])
      ac = apcluster(negDistMat(x1, r=2), q=q)

@@

for (pt in sum.patient.data$Patient) {
  x1 = apclust.data(patient.data[[pt]]$seg.vals, patient.data[[pt]]$info$Samplename)
  if (subset(sum.patient.data, Patient == pt)$Status == 'P' & length(which(grepl('HGD|IMC', patient.data[[pt]]$info$Pathology))) > 0) {
    final = grep('HGD|IMC', patient.data[[pt]]$info$Pathology)
    x1 = apclust.data(patient.data[[pt]]$seg.vals, patient.data[[pt]]$info$Samplename[-final])
  }
  ac = apcluster(negDistMat(x1, r=2), q=q)

@@

ttpc = t.test( subset(simX, Status == 'NP')$clusters, subset(simX, Status == 'P')$clusters )$p.value
ggplot(melt(data.frame(simX[c('Status','clusters')])), aes(x=Status, y=value, fill=Status)) +  
    geom_boxplot() + geom_jitter(shape=16, position=position_jitter(0.2)) +
    stat_boxplot(geom ='errorbar') + labs(y='num. clusters', x='', title=paste('Leave out HGD sample(s) p-value', signif(ttpc, 3)))

fit = glm(Status~clusters, simX, family=binomial(link='logit') )

simX = simX[, c('Status', 'clusters', 'ns','ss','n.var')]
noHGDPR = as.data.frame(matrix(0,1000,2,dimnames=list(c(), c('prec','recall'))))
for (i in 1:nrow(actualMatrix)) {
  trainingRows = sample(1:nrow(simX), nrow(simX)-5)
  train = simX[trainingRows,]
  test = simX[-trainingRows,]

@@

  actual = test$Status == 'P'
  predicted = pred>0

  noHGDPR[i,'prec'] = getPrecision( actual, predicted )
  noHGDPR[i, 'recall'] = getRecall( actual, predicted )
}

#ggplot(as.data.frame(noHGDPR), aes(x=recall, y=prec)) + geom_point() + 
#  geom_hline(yintercept = 0.5, col='red', alpha=0.5) +
#  geom_vline(xintercept = 0.5, col='red', alpha=0.5) +
#  labs(title='Precision/Recall for Progressor/NP prediction NO HGD')

@@

Currently, without the HGD/IMC samples we can predict progressors `r nrow(subset(noHGDPR, prec > 0.5))/nrow(noHGDPR)*100`% of the time.
In comparison, including HGD/IMC samples our predictions improve to `r length(which(actualMatrix[,'prec'] > 0.5))/nrow(actualMatrix)*100`%.

@@

@@

>>>>

@@

patient.info = read.patient.info(patient.file)
patient.info$Patient = gsub("/", "_", patient.info$Patient)
#head(patient.info)

@@

sum.patient.data = summarise.patient.info(patient.info)
sum.patient.data = as.data.frame(subset(sum.patient.data, Patient %in% names(patient.data))) ## For now
pander(sum.patient.data)

# Missing some of the samples as they aren't all sequenced yet
for (pt in names(patient.data)) {
  patient.data[[pt]]$info= patient.data[[pt]]$info[patient.data[[pt]]$info$Samplename %in% colnames(patient.data[[pt]]$seg.vals)[-(1:5)],]
}

@@

plot.pca<-function(pt) {
  print(pt)
  info = arrange(patient.data[[pt]]$info, Endoscopy.Year)
  info = info[which(info$Samplename %in% colnames(patient.data[[pt]][['seg.vals']])[-(1:5)]),]
  plots = list()
  for (sv in c('seg.vals')) {

    x1 = patient.data[[pt]][[sv]][, info$Samplename]
    x1 = x1[,info$Samplename]
    colnames(x1) = paste(info$Endoscopy.Year, '(', info$Pathology, ')', sep='' )

@@

pplots = lapply(subset(sum.patient.data, Status == 'P')$Patient, plot.pca)
#npplots = lapply(subset(sum.patient.data, Status == 'NP')$Patient, plot.pca)
```

```{r echo=F, message=F, warning=F, pcaplots, fig.align='left', fig.height=30, fig.width=20}
grid.arrange(grobs=pplots, nrow=length(subset(sum.patient.data, Status == 'P')$Patient), top="Progressors")
#grid.arrange(grobs=npplots, nrow=length(subset(sum.patient.data, Status == 'NP')$Patient), top="Non-Progressors")
```

```{r persample, eval=F}

lapply(patient.data, function(df) {

  lapply(df$seg.vals[,-(1:5)], quantile)

})

par(mfrow=c(2,2))
  lapply(patient.data$AHM0896$seg.vals[,-(1:5)], hist)

summary(sapply(patient.data$AHM0896$seg.vals[,-(1:5)], mean))
lapply(lapply(patient.data$AHM0896$seg.vals[,-(1:5)], kmeans, 4), function(fit) fit$centers)

km.fits = lapply(patient.data, function(df) kmeans(df$seg.vals[,-(1:5)], 4))
autoplot(km.fits$AHM0896, patient.data$AHM0896$seg.vals[,-(1:5)])

plot(patient.data$AHM0896$seg.vals[,-(1:5)], col=km.fits$AHM0896$cluster, pch=19)

  ## TODO find an OCCAMS patient with genome doubling, downsample and compare

@@

#### EM
```{r clust, echo=F, message=F, warning=F, fig.height=4, fig.width=4, eval=F}
em.dr<-function(patient, status, x1, var.regions) {
  cat(paste('**', patient, '**'))
  clusters=c(2:4)

@@

  for (g in clusters) {
    print(g)
    model = densityMclust(x1, G=g)
    highCluster = which.max(colSums(model$parameters$mean))
    plot(model, what='density')
    # tryCatch({
    #   mod1dr = MclustDR(model, lambda = 0.5)
    #   
    #   #     par(mfrow=c(2,1)) 
    #   plot(mod1dr, what="evalues")
    #   title(paste(pt, '(',status,') G=', g, sep=''))
    #   plot(mod1dr, what='classification') # while this looks like PCA, it's not quite
    #   title(paste(pt, '(',status,') G=', g, sep=''))
    #   #     dev.off()
    #   plot(mod1dr, what='density', dimens=1)
    #   title(paste(pt, '  G=', g, sep=''))
    # }, error = function(e) {
    #   cat(paste(pt, g, "failed DR"))
    # })
    # 
    # a1 = rownames(x1[which(model$classification == highCluster),] )
    # if (nrow(var.regions) > 0) {
    #   am = merge(a1, var.regions, all=T)
    #   pander( as.data.frame(rownames(x1[which(model$classification == highCluster),])), caption=paste("From the cluster with the highest means the regions above are selected, of these ",length(setdiff(rownames(var.regions), a1))," regions are not found in the variable regions selected using cutoffs.", length(intersect(rownames(var.regions), a1)), " are in both."))
    # } else {
    #   pander(as.data.frame(a1),caption=paste("From the cluster with the highest means the regions above are selected. This patient had no variable regions selected based on cutoffs"))
    # 
    # }
    llk[[as.character(g)]] = model
  }
  return(llk)
}

llk = as.data.frame(matrix(nrow=nrow(sum.patient.data),ncol=3,dimnames=list(c(sum.patient.data$Patient), c(2:4))))
for (pt in sum.patient.data$Patient) {
  print(pt)

    status = subset(sum.patient.data, Patient == pt)$Status

@@

```{r kmeansraw, echo=F, message=F, warning=F, eval=F, fig.height=6, fig.width=6, eval=F} 
for (pt in sum.patient.data$Patient) {
    status = subset(sum.patient.data, Patient == pt)$Status
    x1 = patient.data[[pt]]$seg.vals
    rownames(x1) = apply(patient.data[[pt]]$seg.vals[,c(1:4)], 1, function(x) strip.whitespace(paste(x[['chrom']], ':', x[['start.pos']], '-', x[['end.pos']], sep=''))) 
    x1 = x1[,patient.data[[pt]]$info$Samplename]

@@

```{r clust2, echo=F, message=F, warning=F, fig.height=4, fig.width=4, eval=F}
llk = as.data.frame(matrix(nrow=nrow(sum.patient.data),ncol=3,dimnames=list(c(sum.patient.data$Patient), c(2:4))))
for (pt in sum.patient.data$Patient) {
    status = subset(sum.patient.data, Patient == pt)$Status
    x1 = patient.data[[pt]]$norm.seg.vals
    rownames(x1) = apply(patient.data[[pt]]$seg.vals[,c(1:4)], 1, function(x) strip.whitespace(paste(x[['chrom']], ':', x[['start.pos']], '-', x[['end.pos']], sep=''))) 
    x1 = x1[,patient.data[[pt]]$info$Samplename]

@@

# Log liklihood doesn't separate them individually or the range
llk.diff = apply(llk, 1, function(x) diff(range(x)) )
wilcox.test(llk.diff[subset(sum.patient.data, Status == 'NP')$Patient], llk.diff[subset(sum.patient.data, Status == 'P')$Patient])

@@

#### Kmeans
```{r kmeansnorm, echo=F, message=F, warning=F, fig.height=6, fig.width=6, eval=F} 
for (pt in sum.patient.data$Patient) {
    status = subset(sum.patient.data, Patient == pt)$Status
    x1 = patient.data[[pt]]$norm.seg.vals
    rownames(x1) = apply(patient.data[[pt]]$seg.vals[,c(1:4)], 1, function(x) strip.whitespace(paste(x[['chrom']], ':', x[['start.pos']], '-', x[['end.pos']], sep=''))) 
    x1 = x1[,patient.data[[pt]]$info$Samplename]

@@

apclust.data<-function(segdata, samples) {
  x1 = segdata[,intersect(colnames(segdata), samples)]
  #x1 = x1[, samples]
  rownames(x1) = (segdata[,c(1:4)] %>%
    rowwise() %>%
    mutate(location=paste(paste(chrom, arm, sep=''), '.', start.pos, '-', end.pos, sep='')))$location
  return(x1)  
}

@@

```{r apclust, echo=F, message=F, warning=F, fig.height=10, fig.width=10, fig.align='left'}
sample.vals = list()
q=0.2

@@

## Using the segmentation values
sum.patient.data$years = with(sum.patient.data, paste(start.year, end.year, sep='-'))
sim = as.data.frame(sum.patient.data[,c('Status','years')])
rownames(sim) = sum.patient.data$Patient
for (pt in sum.patient.data$Patient) {
  patient.data[[pt]]$info = arrange(patient.data[[pt]]$info,  Pathology, Endoscopy.Year)
  x1 = apclust.data(patient.data[[pt]]$seg.vals, patient.data[[pt]]$info$Samplename)

  gg = ggplot(melt(x1), aes(x=variable, y=value, group=variable, fill=variable)) + 
    geom_boxplot(notch=T, show.legend=F) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position='none') +
    labs(title=paste(pt, patient.data[[pt]]$info$Status[1], sep=' - ')) +   theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

  sample.vals[[pt]] = gg

  info = patient.data[[pt]]$info[which(colnames(x1) %in% patient.data[[pt]]$info$Samplename),]
  colnames(x1) = paste(info$Endoscopy.Year, '(', info$Pathology, ')', sep='' )

@@

  last = ncol(x1)
  if (patient.data[[pt]]$info$Status[1] == 'P' & length(which(grepl('HGD|IMC', info$Pathology)) > 0)  )
    last = sort(grep('HGD|IMC', info$Pathology), decreasing=T)[1]
  if (ncol(x1) <= 15) {
    #plot(ac, x1[,c(c(1:ncol(x1))[-last], last)], main=pt)
  } else {
    #plot(ac, x1[,c(1, sort(sample(c(1:(ncol(x1)-1))[c(-1,-last)], 8)) ,last) ], main=pt)
  }

@@

do.call(grid.arrange, c(sample.vals))

# May be a difference
ttps = wilcox.test( subset(sim, Status == 'NP')$ss, subset(sim, Status == 'P')$ss )$p.value
ttpc = wilcox.test( subset(sim, Status == 'NP')$clusters, subset(sim, Status == 'P')$clusters )$p.value

@@

g2 = ggplot(sim[,c('Status','clusters')], aes(y=clusters, x=Status, fill=Status)) + geom_boxplot() + 
  geom_jitter(width=0.2) + stat_boxplot(geom ='errorbar') +
  labs(title=paste("p-value", signif(ttpc, 3)), y='num. clusters')

sim$Status.Int = as.integer(sim$Status)-1
chisq.test(sim[,c('Status.Int','clusters')])
```

@@

```{r apglmclus, echo=F, message=F, warning=F}
precisionRecall<-function(actual, predicted) {
    retrieved <- sum(predicted==1)
    recall <- sum(predicted==1 & actual==1) / sum(actual==1)
    precision <- sum(predicted==1 & actual==1) / retrieved
    #f1 = signif(2*(1/(1/precision+1/recall)), 2)
    f1 =signif(2*precision*recall/(precision+recall), 2)
    return(data.frame('precision'=precision, 'recall'=recall, 'F1'=signif(f1, 2)))
  }

sim = sim[, c('Status', 'clusters', 'ns','ss')]
actualMatrix = as.data.frame(matrix(0,0,4,dimnames=list(c(), c('status','precision','recall', 'F1'))))
for (i in 1:100) {
  trainingRows = sample(1:nrow(sim), nrow(sim)-10)
  train = sim[trainingRows,]
  test = sim[-trainingRows,]

@@

  #glm_link_scores <- predict(fit, test, type="link")
  #glm_response_scores <- predict(fit, test, type="response")

  actualMatrix = rbind(actualMatrix, cbind('status'='P', precisionRecall(test$Status == 'P', pred>0)))
  actualMatrix = rbind(actualMatrix, cbind('status'='NP', precisionRecall(test$Status == 'NP', pred<0)))
}

g1 = ggplot(actualMatrix, aes(x=recall, y=precision, color=status, label=F1)) + 
  geom_point() + geom_text(aes(label=ifelse(precision>0.5 & recall>0.5, as.character(F1), ''), hjust=0, vjust=0)) +
   geom_hline(yintercept = 0.5, col='grey') + geom_vline(xintercept = 0.5, col='grey') +
   labs(title='Precision/Recall for Progressor/NP prediction')

g2 = ggplot(actualMatrix, aes(x=status, y=F1, group=status, fill=status)) + 
  geom_violin() + geom_jitter(alpha=0.5) + ylim(0,1) + 
  geom_violin(data=subset(actualMatrix,precision>0.5 & recall>0.5), aes(x=status, y=F1, group=status, fill='P&R>0.5') , alpha=0.6) +
  labs(title="F1 statistic")

grid.arrange(g1, g2, top='100 iterations 10f xval')
```

Randomize the data and see if the F1 statistic still holds

```{r gpglmclus_rand, echo=F, message=F, warning=F}
sim = sim[, c('Status', 'clusters')]

rand = as.data.frame(matrix(0,0,4,dimnames=list(c(), c('status','precision','recall', 'F1'))))
for (i in 1:100) {
  # rand.sim = as.data.frame(cbind(  'Status'=sample(c('P','NP'), nrow(sim), replace=T, prob=round(table(sim$Status)/nrow(sim), 2)),
  #                    'clusters'=sample((3:30), nrow(sim), replace=T) ))
  # rand.sim = as.data.frame(cbind(  'Status'=sample(c('P','NP'), nrow(sim), replace=T, prob=round(table(sim$Status)/nrow(sim), 2)),
  #                    'clusters'=sample((3:30), nrow(sim), replace=T) ))

  rand.sim = as.data.frame(cbind(  'Status'=sample(c('P','NP'), nrow(sim), replace=T),
                     'clusters'=sim$clusters ))

  rand.sim$clusters = as.integer(as.character(rand.sim$clusters))

  trainingRows = sample(1:nrow(rand.sim), nrow(rand.sim)-10)
  train = rand.sim[trainingRows,]
  test = rand.sim[-trainingRows,]

  rand = rbind(rand, cbind('status'='P', precisionRecall(test$Status == 'P', pred>0)))
  rand = rbind(rand, cbind('status'='NP', precisionRecall(test$Status == 'NP', pred<0)))
}

g1 = ggplot(rand, aes(x=recall, y=precision, color=status, label=F1)) + 
  geom_point() + geom_text(aes(label=ifelse(precision>0.5 & recall>0.5, as.character(F1), ''), hjust=0, vjust=0)) +
   geom_hline(yintercept = 0.5, col='grey') + geom_vline(xintercept = 0.5, col='grey') +
   labs(title='Precision/Recall for Progressor/NP prediction')

g2 = ggplot(rand, aes(x=status, y=F1, group=status, fill=status)) + 
  geom_violin() + geom_jitter(alpha=0.5) + ylim(0,1) + 
  #geom_violin(data=subset(actualMatrix,precision>0.5 & recall>0.5), aes(x=status, y=F1, group=status, fill='P&R>0.5') , alpha=0.6) +
  labs(title="F1 statistic")

grid.arrange(g1, g2, top='100 iterations 10f xval - randomized labels')

@@

cols = c('chr','cluster', 'P.x','x')
hgmx = matrix(nrow=0,ncol=length(cols),dimnames=list(c(), cols))
for (chr in c(1:22)) {
  for (i in 1:length(ac@clusters)) {
    cl = x1[ac@clusters[[i]],]

@@

```{r hgmx, echo=F, message=F, warning=F, fig.height=6, fig.width=6}

g1 = ggplot(subset(hgmx, x > 0), aes(x=factor(chr),y=P.x,group=chr,fill=factor(chr))) + ylim(0,0.5)+ 
  geom_boxplot() +
  labs(title='Probability of cluster assignment', x='Chromosome', y='P(x)')

g2 = ggplot(subset(hgmx, x > 0), aes(x=factor(cluster),y=P.x,group=cluster,fill=factor(cluster))) + ylim(0,0.5)+ 
  geom_boxplot() +
  labs(title='Probability of cluster assignment', x='Cluster', y='P(x)')

grid.arrange(g1, g2, nrow=2)
```

@@

### Variable Regions

Should we call differently?  One is regions that are not diploid, the second would be what changes between samples?

```{r varregions, echo=F, message=F, warning=F, eval=F}
pt = 'AHM0896'

se = round(apply(patient.data[[pt]]$seg.vals[,-(1:5)], 2, function(x) sd(x)/sqrt(length(x))), 3)
sdx = round(apply(patient.data[[pt]]$seg.vals[,-(1:5)], 2, sd), 3)

row=325
patient.data[[pt]]$seg.vals[row,]
round(patient.data[[pt]]$seg.vals[,-(1:5)][row,] + sdx, 1)
round(patient.data[[pt]]$seg.vals[,-(1:5)][row,] - sdx, 1)

lapply(patient.data, function(df) nrow(df$variable.regions))

@@

```{r xval_pt, echo=F, warning=F, message=F, fig.align='left', fig.height=4, fig.width=4, eval=F}
# Leave out patients
loo_xval = matrix(ncol=3, nrow=0, dimnames=list(c(), c('net.similarity', 'sum.similarity', 'clusters')))
for (i in 1:nrow(sum.patient.data)) {
  simL = sum.patient.data[-i,c(2:4)]
  rownames(simL) = sum.patient.data[-i,1]

@@

  for (pt in sum.patient.data$Patient) {
      if (length(patient.data[[pt]]$info$Samplename) <= 2) next
      x1 = apclust.data(patient.data[[pt]]$seg.vals, patient.data[[pt]]$info$Samplename[-sample(1:(length(patient.data[[pt]]$info$Samplename)), 1)])
      ac = apcluster(negDistMat(x1, r=2), q=q)

@@

for (pt in sum.patient.data$Patient) {
  x1 = apclust.data(patient.data[[pt]]$seg.vals, patient.data[[pt]]$info$Samplename)
  if (subset(sum.patient.data, Patient == pt)$Status == 'P' & length(which(grepl('HGD|IMC', patient.data[[pt]]$info$Pathology))) > 0) {
    final = grep('HGD|IMC', patient.data[[pt]]$info$Pathology)

    if (length(patient.data[[pt]]$info$Samplename[-final]) <= 1) next

    x1 = apclust.data(patient.data[[pt]]$seg.vals, patient.data[[pt]]$info$Samplename[-final])
  }
  ac = apcluster(negDistMat(x1, r=2), q=q)

@@

ct = cor.test(simX$clusters, simX$n.var)

ttpc = t.test( subset(simX, Status == 'NP')$clusters, 
               subset(simX, Status == 'P')$clusters )$p.value

ggplot(melt(data.frame(simX[c('Status','clusters')])), aes(x=Status, y=value, fill=Status)) +  
    geom_boxplot() + geom_jitter(shape=16, position=position_jitter(0.2)) +
    stat_boxplot(geom ='errorbar') + labs(y='num. clusters', x='', title=paste('Leave out HGD sample(s) p-value', signif(ttpc, 3)))

#fit = glm(Status~clusters, simX, family=binomial(link='logit') )

simX = simX[, c('Status', 'clusters', 'ns','ss','n.var')]
noHGDPR = as.data.frame(matrix(0,0,4,dimnames=list(c(), c('status','precision','recall', 'F1'))))
for (i in 1:100) {
  trainingRows = sample(1:nrow(simX), nrow(simX)-10)
  train = simX[trainingRows,]
  test = simX[-trainingRows,]

@@

  noHGDPR = rbind(noHGDPR, cbind('status'='P', precisionRecall(test$Status == 'P', pred>0)))
  noHGDPR = rbind(noHGDPR, cbind('status'='NP', precisionRecall(test$Status == 'NP', pred<0)))

}

g1 = ggplot(noHGDPR, aes(x=recall, y=precision, color=status, label=F1)) + 
  geom_point() + geom_text(aes(label=ifelse(precision>0.5 & recall>0.5, as.character(F1), ''), hjust=0, vjust=0)) +
   geom_hline(yintercept = 0.5, col='grey') + geom_vline(xintercept = 0.5, col='grey') +
   labs(title='Precision/Recall for Progressor/NP prediction without HGD')

g2 = ggplot(subset(actualMatrix, precision>0.5 & recall>0.5), aes(x=status, y=F1, group=status, fill=status)) + 
  geom_violin() + geom_jitter(alpha=0.5) + ylim(0.5,1) + labs(title="F1 statistic at prec+recall>0.5")

grid.arrange(g1, g2, top='100 iterations 10f xval no HGD')

@@

Currently, without the HGD/IMC samples we can predict progressors `r nrow(subset(noHGDPR, precision > 0.5))/nrow(noHGDPR)*100`% of the time.
In comparison, including HGD/IMC samples our predictions improve to `r length(which(actualMatrix[,'precision'] > 0.5))/nrow(actualMatrix)*100`%.  It's also worth noting that the number of clusters has a significant (pval=`r signif(ct$p.value, 1)`) correlation to the number of variant regions identified across all samples for that patient (r=`r round(ct$estimate, 2)`).

@@

```{r}

pts = rownames(subset(simX, Status == 'P' & clusters <= mean(subset(simX, Status == 'NP')$clusters)))

subset(sum.patient.data, Patient %in% pts)

@@

## Investigating Cutoff selection

Currently we use the same cutoffs across all patients and samples for selecting variable regions. While it's clear based on the data that values ~1 are the average ploidy of the sample and are most likely to be diploid, some samples have a greater spread round 1 resulting in the possibility that we are both under and over calling variable regions.

In these plots there are several things occurring.  The values have been log-10 transformed in order to spread them out. These are then plotted in a histogram with two sets of lines representing potential cutoffs: one for EM derived means, the other for mean(+|-)sd*1.5.  In most cases EM does not provide much separation. This could potentially be improved by adding priors or other parameters to the model, but it's unclear how to initialize these usefully.

```{r varcall, echo=F, message=F, warning=F, fig.height=10, fig.width=10}
for (pt in names(patient.data)) {
  cols = intersect(arrange(patient.data[[pt]]$info, Pathology, Endoscopy.Year)$Samplename, colnames(patient.data[[pt]]$seg.vals[,-(1:5)]))
  segvals = patient.data[[pt]]$seg.vals[,cols]

  plot.seg.vals<-function(lt, title, hist.col='blue') {
    sdx = apply(lt, 2, sd)
    se = apply(lt, 2, function(x) sd(x)/sqrt(length(x)))
    means = t(do.call(rbind, lapply(lapply(lt, densityMclust, G=3), function(x) sort(x$parameters$mean))))
    m = merge(melt(lt), melt(means), by.x='variable', by.y='Var2')
    gg = ggplot(m,aes(value.x, ..density..)) + 
      facet_wrap(~variable,scales = "free_x") + 
        geom_histogram(bins=10, fill=hist.col, color='grey') +
        geom_density(aes(value.x), color='darkgrey', linetype='dotted') +
        geom_vline( aes(xintercept=value.y, col='EM'), linetype='twodash') +
        geom_vline(aes(xintercept = mean(value.x)+sd(value.x)*1.5, color='StdDev'), linetype='twodash') +
        geom_vline(aes(xintercept = mean(value.x)-sd(value.x)*1.5, color='StdDev'),linetype='twodash') +
        labs(title=title)
    return(gg)
  }

  gg1 = plot.seg.vals(segvals, title=paste(pt, ' (',patient.data[[pt]]$info$Status[1],')', sep=''))
  gg2 = plot.seg.vals(log2(segvals), title=paste(pt, ' log2  (',patient.data[[pt]]$info$Status[1],')', sep=''), hist.col='darkgreen')
  gg3 = plot.seg.vals(log10(segvals), title=paste(pt, ' log10  (',patient.data[[pt]]$info$Status[1],')', sep=''), hist.col='darkred')

  print(gg1)
  print(gg2)
  print(gg3)
  #do.call(grid.arrange, list(gg1, gg2, gg3))

}
```
